Wait-free Queues with Polylogarithmic Step Complexity

### 2022-10-11 thesis/thesis.tex
update double refresh fig
apply Oct 11 notes

\paragraph{Introduce tree} There is a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes. Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as linearization ordering. 
\begin{figure}[hbt]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{\label{fig::blocktree}Each of the processes $P_1,P_2...P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}
\it{TODO::Add sequence to nodes}

\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, and then tries to append them to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh()}. A process doing a \nf{Refresh} on $n$ with successful append helps other processes doing \nf{Refresh} on $n$ concurrently to propagate their operations up to the parent. The key idea is that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in the node after the second failed \tt{Refresh}.
This is because if both of the \tt{Refresh}es on $n$ fail to append then there is another instance of \nf{Refresh} in between which has gathered its new operations after the first failed \nf{Refresh} and succeeded to do an append. This \nf{Refresh} appends the operations that the first failed \nf{Refresh} was trying to append.

\begin{figure}[hbt]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{New operations are appended.}
\end{subfigure}
\hspace{6em}
\caption{\label{fig::propagstep} Before and after of a \nf{$n$.Refresh} with successful append. Operations propagating from the left child are numbered with $l$ and from the right child by $r$.}
\end{center}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent with a successful \texttt{Refresh}.}
\end{figure}
\it{TODO::simplify the Figure, remove code invocations}

@@

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}

@@

\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would no be poly-logarithmic time.
Instead we use implicit representation of the operations propagated together. Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}). Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}, a piece of information about the operations in a \nf{Refresh} step. A block contain the number of operations from the left and the right child in a \texttt{Refresh} procedure. A node stores an array of blocks of operations propagted up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Furthermore, since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

\begin{figure}
\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]

\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]

\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Showing concurrent operation sets with blocks. Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively. Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child. Blocks between two lines $||$ are propagated together to the parent. For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent. The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to. Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$. For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child. Block (5,3) also points to elements from blocks (0,5) and (1,2). We choose to linearize operations in a block from the left child before those from the right child as a convention. Operations within a block of the root can be ordered in any way that is convenient. In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids. So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2). All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
}
\end{figure}

@@

\paragraph{Explain pieces of a block helping to do Get and Index}
After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.

To get the $i$th enqueue in the root, we can find the block $B$ containing $i$th element in the root, and then recursively find the subblock of $B$ that contains $i$th element. To make this recursive search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search. We also store pointers to determine the range of subblocks of a block to make binary search faster. In each block, we store the prefix sum of operations from the left child and the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figures \ref{fig:prefix} and \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

\begin{figure}
\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]

\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]

\end{tikzpicture}
\end{center}
\caption{\label{fig:prefix} Using Prefix sums in blocks. When we want to find block b elements in its children, we can use binary search. The number below each block shows the count of elements in the previous blocks.}
\end{figure}

\begin{figure}[hbt]
\centering
  \includegraphics[width=5in]{pics/pointers}
  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
\end{figure}

\paragraph{Introduce how to create Queue using Get and Index} In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a \nf{Dequeue} and one tries to get an \nf{Enqueue} with an specific rank. \nf{Enqueue}s and \nf{Dequeue}s are appended to the tree and when we want to find the response to a \nf{Dequeue}, we compute the place of the \nf{dequeue} in the linearization. Then, using the rank of the \nf{Dequeue} among \nf{Deqeueue}s and the information about the size of the queue stored in the root we compute which \nf{Enqueue} is the answer to the \nf{Dequeue} or if the answer is null.

\paragraph{}
A non-null \nf{Dequeue} is one that returns a non-null value.
If the queue is non-empty and $r$ \texttt{Dequeue} operations have returned a non-null response, then $i$th \texttt{Dequeue} returns the input of the $r+1$th \texttt{Enqueue}. So in order to answer a \nf{Dequeue}, it iss sufficient to know whether the queue is empty or not and the number of previous non-null \nf{dequeue}s.

\paragraph{}
We know we can linearize operations in a block in any order; here, we choose to decide to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. In the next example, operations in a cell are concurrent. \texttt{Dequeue()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively. We will next describe how these values can be computed efficiently.

\begin{table}[hbt]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
  \end{tabular}
\end{small}

@@

\begin{table}[hbt]
\centering
\begin{small}
  \begin{tabular}{c|c|c|c|c}
    \hline &\texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
    \tt{\#ENQs} & 0 & 3 & 1 & 1 \\ \hline
    \tt{\#DEQs} & 1 & 1 & 1 & 4 \\ \hline
    \tt{size} & 0 & 2 & 2 & 0 \\ \hline
  \end{tabular}
\end{small}
  \caption{Augmented history of operation blocks on the queue}
\end{table}

@@

Given \texttt{DEQ} is in block $B$, response of \nf{DEQ}  is the \tt{Enqueue} with index \nf{non-null Dequeus[b-1]+ index of DEQ in the block's Dequeus} in the root, if \texttt{(size[b-1]- index of DEQ in the $B$'s Dequeus >=0)}. Otherwise it would be \nf{ null}

\begin{figure}[hbt]
\centering
  \includegraphics[width=5in]{pics/queue}
  \caption{Fields stored in the Queue nodes. \label{fig::queue}}
\end{figure}

@@

\paragraph{Block}
Block $b$ contains subblocks in the left and right children. The left subblocks of $b$ are some consecutive blocks in the left child starting from where the block prior to $b$ ended. See Figure \ref{fig::createBlock} .

We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s. In each \texttt{node} we store pointers to its parent and children, an array of \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}. Furthermore, in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}. There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh} step. It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases. Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}. This helps us to know which blocks have aggregated together to a block, not precisely though. We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.

In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{num\textsubscript{enq-left}}, \texttt{num\textsubscript{deq-left}}, \texttt{num\textsubscript{enq-right}}, \texttt{num\textsubscript{deq-right}}. In order to make \texttt{BinarySearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.

To compute the head of the \texttt{queue} before a \texttt{dequeue} two more fields are stored in the root \texttt{size} and \texttt{sum\textsubscript{non-null deq}}. \texttt{size} in a \texttt{block} shows the number of elements after the \texttt{block} has finished and \texttt{sum\textsubscript{non-null deq}} is the total number of non-null dequeues till the \texttt{block}.

@@

\it{I believe the code is readable and we do not need references to code.}

@@

\paragraph{FindResponse(b,i)}
To compute the response of the $i$th \nf{dequeue} in $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not. If there are more dequeues than enqueues the queue would become empty before the requeusted dequeue. If the queue was not empty, Line \ref{ki} 222 computes the position of the eqneueue which is response to the dqeueue. Knowing the response is the $e$th enqueue in the root which is before the $b$th block we find the block and position contaiing the qnueue operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.

@@

\paragraph{n.Refresh()}
\nf{$n$.Refresh} goal is to  create a block of $n$'s children new blocks and append it to $n$\nf{.blocks}. From Lines 311 to 316 \ref{li} \nf{$n$.Refresh} helps to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} and set the \nf{super} field of the last block in $n$. The reason behind helping \nf{super} field is explained later in \nf{IndexDequeue} paragraph. After helping to \nf{Advance} children, the value $h$ from $n$\nf{.head} is read at Line \ref{readHead} and \nf{new} block is created in Line \ref{invokeCreateBlock}. \nf{new} is going to be inserted into $n\nf{.block[h]}$. The reason why reading \nf{head} and \nf{CreateBlock} are in two lines is that maybe the \nf{head} value changes from reading it till creating a block to be inserted to \nf{head}. Then if \nf{new} is empty \nf{Refresh} returns \nf{true} because there is no new operations to propagate  an it is unnecessary to add an empty block to the tree. Later we will assume all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}, if it was succsuful all is good, if not it means someone else has put a block earlier into \nf{head} or maybe the \nf{head} was not empty at first point. Anyway \nf{Refresh} tries to update the \nf{head} and \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

@@

\it{should I get into details of computing size and sumenq?}

\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/createblock}
  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
\end{figure}

\paragraph{DSearch(e,end)}
We can show an operation in a node in two formats, the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block. If we know tuple of block and rank in the block of an operation ($E$) we can find the this tuple in the subblock containig the operation in poly log time. To find the response of a dequeue, the information we know about the response enquue is like the first way of showing the operation in the root ($E_i(\nf{root})$). So we have to find the block in the root containing the \nf{e}th enqueue in the root. We also know \nf{end} which the \nf{e}th enqueu is before or in the \nf{root.blocks[end]}. \nf{DSearch} uses the Doubling idea to find the range that the answer block is there (Lines \ref{dsearchStart}-\ref{dsearchEnd}) and then tries to find the response with binary search (Line 806\ref{lo}).

@@

\Function{boolean}{Refresh()}{}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{h= \head} \label{readHead}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}

>>>>

\paragraph{Introduce tree} There is a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes. Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{\label{fig::blocktree}Each of the processes $P_1,P_2...P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh()}. A process doing a \nf{Refresh} on $n$ with successful append helps other processes doing \nf{Refresh} on $n$ concurrently to propagate their operations up to the parent. The key idea is that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in the node after the second failed \tt{Refresh}.
This is because if both of the \tt{Refresh}es on $n$ fail to append then there is another instance of \nf{Refresh} in between which has gathered its new operations after the first failed \nf{Refresh} and succeeded to do an append. This \nf{Refresh} appends the operations that the first failed \nf{Refresh} was trying to append.

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{New operations are appended.}
\end{subfigure}
\hspace{6em}
\caption{\label{fig::propagstep} Before and after of a \nf{$n$.Refresh} with successful append. Operations propagating from the left child are numbered with $l$ and from the right child by $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
  \caption{The second failed Refreshof $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its operations after the first failed \nf{Refresh}.}
\end{figure}

@@

\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would no be poly-logarithmic time.
Instead we use implicit representation of the operations propagated together. Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}). Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}, a piece of information about the operations in a \nf{Refresh} step. A block contain the number of operations from the left and the right child in a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example). A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Furthermore, since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]

@@

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture}
[level 1/.style={level distance=3.3cm,sibling distance=0cm},
	level 2/.style={level distance=2.5cm,sibling distance=0cm},
	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
\begin{small}

\Tree [.{\begin{tabular}{||c|c|c||}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline (2,1) & (5,0) & (4,2) & (3,4) & (6,3) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Showing concurrent operation sets with blocks. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively. For example, block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child. Block (5,3) is superblock of blocks (0,5) and (1,2) and blocks (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

@@

\paragraph{Explain pieces of a block helping to do Get and Index}
After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). Each block stores an extra constant amount of information (like prefix sums) to allow binary search to find the required block in a node quickly

To get the $i$th operation in the root, we find the block $B$ containing $i$th operation in the root, and then recursively find the subblock of $B$ that contains $i$th element. To make this recursive search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search. We also store pointers to determine the range of subblocks of a block to make binary search faster. In each block, we store the prefix sum of operations from the left child and the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in. After a block is installed we set the approximate index of the superblock of the block in it.

\begin{figure}[hbtp]
\centering
  \includegraphics[width=4in, height=1.4in]{pics/pointers}
  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
\end{figure}

\paragraph{Introduce how to create Queue using Get and Index}

In an execution on a queue where no dequeue operation returns null, the $k$th dequeue returns the element of the $k$th enqueue. In general case a dequeue returns null if and only if the size of the queue after the previous operation is 0. Otherwise if the the dequeue is the $k$th non null returning dequeue it returns the element of $k$th enqueue. Having the size of the  queue after an operation we can compute the number of non null returning dequeues  from the number of enqueues up to block B and size of the queue after B. So if we store the size of the queue after each block operations in the root we can compute the index of the enqueue which is response to a dequeue in constant time.

In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a \nf{Dequeue} and one tries to get an \nf{Enqueue} with an specific rank.  We know we can linearize operations in a block in any order; here, we choose to decide to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. In the next example, operations in a cell are concurrent. \texttt{Dequeue()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.

\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
  \end{tabular}
\end{small}

@@

\begin{table}[hbtp]
\centering
\begin{small}
  \begin{tabular}{c|c|c|c|c}
    \hline &\texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
    \tt{\#ENQs} & 0 & 3 & 1 & 1 \\ \hline
    \tt{\#DEQs} & 1 & 1 & 1 & 4 \\ \hline
    \tt{size} & 0 & 2 & 2 & 0 \\ \hline
  \end{tabular}
\end{small}
  \caption{Augmented history of operation blocks on the queue}
\end{table}

\it{Do you think the remainig part is overkill? I think it is to much detail which is also repeated in the proof.}

@@

Given \texttt{DEQ} is in block $B$, response of \nf{DEQ}  is the \tt{Enqueue} with index \nf{non-null Dequeus[b-1]+ index of DEQ in the block's Dequeus} in the root, if \texttt{(size[b-1]- index of DEQ in the $B$'s Dequeus >=0)}. Otherwise it would be \nf{null}

@@

\paragraph{Node} (\it{Repetitive as the fields description?})
 In each \texttt{node} we store pointers to its parent and children, an array of \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{Block} (\it{Repetitive as the fields description?})
Block $b$ contains subblocks in the left and right children. The left subblocks of $b$ are some consecutive blocks in the left child starting from where the block prior to $b$ ended. In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. \nf{end\sub{left}} and \nf{end\sub{right}} field show the last subblock of a block in left and right child.  \texttt{size} in a \texttt{block} in \nf{root} stores the size of the queue after the \texttt{block} has finished.

@@

\it{I believe the code is readable enough and we do not need references to code.}

@@

\paragraph{FindResponse(b,i)}
To compute the response of the $i$th \nf{dequeue} in $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not. If there are more dequeues than enqueues the queue would become empty before the requeusted dequeue. If the queue was not empty, Line \ref{computeE} computes the position of the eqneueue which is response to the dqeueue. Knowing the response is the $e$th enqueue in the root which is before the $b$th block we find the block and position contaiing the qnueue operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.

@@

\paragraph{n.Refresh()}
\nf{$n$.Refresh} goal is to  create a block of $n$'s children new blocks and append it to $n$\nf{.blocks}. Lines \ref{startHelpChild1}-\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} and set the \nf{super} field of the last block in $n$. The reason behind helping \nf{super} field is explained later in \nf{IndexDequeue} paragraph. After helping to \nf{Advance} children, the value $h$ from $n$\nf{.head} is read at Line \ref{readHead} and \nf{new} block is created in Line \ref{invokeCreateBlock}. \nf{new} is going to be inserted into $n\nf{.block[h]}$. The reason why reading \nf{head} and \nf{CreateBlock} are in two lines is that maybe the \nf{head} value changes from reading it till creating a block to be inserted to \nf{head}. Then if \nf{new} is empty \nf{Refresh} returns \nf{true} because there is no new operations to propagate  an it is unnecessary to add an empty block to the tree. Later we will assume all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}, if it was succsuful all is good, if not it means someone else has put a block earlier into \nf{head} or maybe the \nf{head} was not empty at first point. Anyway \nf{Refresh} tries to update the \nf{head} and \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

@@

\it{Should I get into details of computing size and sumenq?}

%\begin{figure}[hbt]
%  \center\includegraphics[width=5.5in]{pics/createblock}
%  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
%\end{figure}

\paragraph{DSearch(e,end)}
We can show an operation in a node in two formats, the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block. If we know tuple of block and rank in the block of an operation ($E$) we can find the this tuple in the subblock containig the operation in poly log time. To find the response of a dequeue, the information we know about the response enquue is like the first way of showing the operation in the root ($E_i(\nf{root})$). So we have to find the block in the root containing the \nf{e}th enqueue in the root. We also know \nf{end} which the \nf{e}th enqueu is before or in the \nf{root.blocks[end]}. \nf{DSearch} uses the Doubling idea to find the range that the answer block is there (Lines \ref{dsearchStart}-\ref{dsearchEnd}) and then tries to find the response with binary search (Line \ref{dsearchBinarySearch}).

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}
